\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Model

### 2021-01-14 Tex/accepted_preprint/accepted.tex
updated accepted preprint files

\documentclass{cup-pan}
\usepackage[utf8]{inputenc}
\usepackage{blindtext}
\usepackage{bm}
%\usepackage{lmodern}
%\usepackage{amsthm}
%\usepackage{geometry} 
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}

% original loaded packages
%\documentclass{cup-pan}
%\usepackage[utf8]{inputenc}
%\usepackage{blindtext}
%\usepackage{bm}
%\usepackage[T1]{fontenc}
%\usepackage{lmodern}

\title{Generative Dynamics of Supreme Court Citations: Analysis with a New Statistical Network Model}

\author[1]{Christian S. Schmid}
\author[2]{Ted Hsuan Yun Chen}
\author[3]{Bruce A. Desmarais}

\affil[1]{Department of Statistics, The Pennsylvania State University, University Park, PA. Email: \url{schmid@psu.edu}}
\affil[2]{Faculty of Social Sciences, University of Helsinki, Finland. Email: \url{ted.hsuanyun.chen@gmail.com}}
\affil[3]{Department of Political Science, The Pennsylvania State University, University Park, PA. Email: \url{bdesmarais@psu.edu}}

%% Corresponding author
\corrauthor{Christian S. Schmid}

%% Abbreviated author list for the running footer
\runningauthor{Schmid et al.}

\addbibresource{bib.bib}

@@

\maketitle

\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part from opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citations are treated as if they arise independently. We present a methodology for studying citations between Supreme Court opinions at the dyadic level, as a network, that overcomes these limitations. This methodology---the citation exponential random graph model, for which we provide user-friendly software---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We then analyze a network that includes all Supreme Court cases decided between 1950 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, and popularity. The dependence effects are as substantively and statistically significant as the effects of exogenous covariates, indicating that models of Supreme Court citation should incorporate both the effects of case characteristics and the structure of past citations.

\keywords{US Supreme Court, majority opinions, citations, network, ERGM.}
\end{abstract}

\section{Introduction}
\label{sec:overview}

@@

\begin{figure}[bt]
\centering
\includegraphics[width = 0.49\textwidth,trim= 2cm 2cm 3cm 2cm,clip=true ]{citations_trans.pdf}\hspace{0.2cm}
\includegraphics[width = 0.49\textwidth,trim= 2cm 2cm 3cm 2cm,clip=true ]{citations_recip.pdf}
\caption{Illustrations of transitive triangle connecting US Supreme Court opinions through citations (left) and a reciprocal tie between two US Supreme Court opinions (right).}
\label{fig:transitivity}
\end{figure}

@@

The third, and final, network property we consider in the context of Supreme Court citations is popularity. Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence,chayes2013mathematics}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed. and for which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. The language in this passage suggests that the attention to Baker v. Carr in previous Court opinions is in part responsible for its authority in Oregon v. Mitchell. The citations to Baker v. Carr are visualized in Figrue \ref{fig:popularity}.
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 time the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

@@

\begin{table}[bt]
\caption{For the time range of interest (1937 - 2015) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** J Roberts still serves as chief justice (retrieved 5/2020).}
\label{tab:chiefs}
\centering
\begin{tabular}{l l l l r}
\headrow \thead{Chief Justice} & \thead{Terms} & \thead{Total Number Cases} & \thead{Cases/Term}  \\
CE Hughes*              & 1937 - 1941                                          & 628                                                               & 125.6                                                     \\ 
HF Stone                & 1942 - 1946                                          & 756                                                               & 151.2                                                     \\ 
FM Vinson               & 1946 - 1953                                          & 789                                                               & 98.63                                                      \\ 
E Warren                & 1954 - 1969                                          & 2149                                                              & 126.41                                                     \\ 
WE Burger               & 1970 - 1986                                          & 2805                                                             & 155.83                                                     \\ 
W Rehnquist            & 1987 - 2005                                          & 2022                                                              & 106.42                                                      \\ 
J Roberts **           & 2006 - 2015                                          & 871                                                              & 87.1                                           \\ 
\end{tabular}

\end{table}

@@

\begin{table}[bt]
\caption{Assigned numbers for the variable \textit{Issue Area}. This information originates from the Supreme Court Database code book.}
\label{issue_area_coding}
\centering
\begin{tabular}{llll}
1 & Criminal Procedure & 8  & Economic Activity    \\
2 & Civil Rights       & 9  & Judicial Power       \\
3 & First Amendment    & 10 & Federalism           \\
4 & Due Process        & 11 & Interstate Relations \\
5 & Privacy            & 12 & Federal Taxation     \\
6 & Attorneys          & 13 & Miscellaneous        \\
7 & Unions             & 14 & Private Action      
\end{tabular}
\end{table}

@@

\section*{Funding}
This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219.

\section*{Acknowledgements}
We thank Rachel Hinkle, Lauren Santoro, and Michael Nelson for helpful feedback on this project.

\section*{Data Availability Statement}
The replication code and data for this paper can be found at \citet{replicationdata}.

\printbibliography
\end{document}

>>>>

\documentclass[headsepline=true, abstracton]{scrartcl}

\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bm}

\usepackage{natbib}
\usepackage{hyperref}
\hypersetup{colorlinks=true, citecolor=darkgray, linkcolor=darkgray, urlcolor=darkgray}

\usepackage[moderate]{savetrees}

\usepackage[table,xcdraw]{xcolor}

\usepackage{graphicx}

 \usepackage{float}

\usepackage{setspace}

\usepackage{url}

\usepackage{geometry}

@@

\renewcommand{\refname}{Bibliography}

\onehalfspacing
\setlength{\headsep}{15mm}

\thispagestyle{plain}

\title{\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Network Model\thanks{
		This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219. We thank Rachel Hinkle, Benjamin Kassow, Lauren Santoro, and Michael Nelson for helpful feedback on this project. The replication code and data for this paper can be found at \citet{replicationdata}.}}
%\author{}
 % Toggle % below to blind and unblind the manuscript
%  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
%   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
% \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
% \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
% \affil[2]{Department of Political Science, Pennsylvania State University}
 %\affil[1]{Department of Statistics, Pennsylvania State University}

\author{%
  Christian S. Schmid\footnote{Department of Statistics, The Pennsylvania State University, schmid@psu.edu}%
  \and Ted Hsuan Yun Chen\footnote{Faculty of Social Sciences, University of Helsinki, ted.hsuanyun.chen@gmail.com}%
   \and Bruce A. Desmarais\footnote{Department of Political Science, The Pennsylvania State University, bdesmarais@psu.edu}%
%  \and David R. Hunter \footnote{Department of Statistics, The Pennsylvania State University, dhunter@stata.psu.edu}%
  }

\date{Conditionally Accepted at \textit{Political Analysis}\\ December 29, 2020}

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part from opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citations are treated as if they arise independently. We present a methodology for studying citations between Supreme Court opinions at the dyadic level, as a network, that overcomes these limitations. This methodology---the citation exponential random graph model, for which we provide user-friendly software---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We then analyze a network that includes all Supreme Court cases decided between 1950 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, and popularity. The dependence effects are as substantively and statistically significant as the effects of exogenous covariates, indicating that models of Supreme Court citation should incorporate both the effects of case characteristics and the structure of past citations.
\end{abstract}

\onehalfspacing
\section{Introduction}
\label{sec:overview}

@@

\begin{figure}[bt]
	\centering
	\includegraphics[width = 0.49\textwidth,trim= 2cm 2cm 3cm 2cm,clip=true ]{citations_trans.pdf}\hspace{0.2cm}%
	\includegraphics[width = 0.49\textwidth,trim= 2cm 2cm 3cm 2cm,clip=true ]{citations_recip.pdf}
	\caption{Illustrations of transitive triangle connecting US Supreme Court opinions through citations (left) and a reciprocal tie between two US Supreme Court opinions (right).}
	\label{fig:transitivity}
\end{figure}

@@

The third, and final, network property we consider in the context of Supreme Court citations is popularity. Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence,chayes2013mathematics}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed. and for which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. The language in this passage suggests that the attention to Baker v. Carr in previous Court opinions is in part responsible for its authority in Oregon v. Mitchell. The citations to Baker v. Carr are visualized in Figure \ref{fig:popularity}.
\begin{quotation}
	The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 time the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

@@

\begin{table}[bt]
	\label{tab:chiefs}
	\centering
	\begin{tabular}{l l c r}
		 {Chief Justice} & {Terms} & {Total Number Cases} & {Cases/Term} \\ \hline
		CE Hughes*              & 1937 - 1941                                          & 628                                                               & 125.6                                                     \\ 
		HF Stone                & 1942 - 1946                                          & 756                                                               & 151.2                                                     \\ 
		FM Vinson               & 1946 - 1953                                          & 789                                                               & 98.63                                                      \\ 
		E Warren                & 1954 - 1969                                          & 2149                                                              & 126.41                                                     \\ 
		WE Burger               & 1970 - 1986                                          & 2805                                                             & 155.83                                                     \\ 
		W Rehnquist            & 1987 - 2005                                          & 2022                                                              & 106.42                                                      \\ 
		J Roberts **           & 2006 - 2015                                          & 871                                                              & 87.1                                           \\ 
	\end{tabular}
		\caption{For the time range of interest (1937 - 2015) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** J Roberts still serves as chief justice (retrieved 5/2020).}
\end{table}

@@

\begin{table}[bt]

	\label{issue_area_coding}
	\centering
	\begin{tabular}{llll}
		1 & Criminal Procedure & 8  & Economic Activity    \\
		2 & Civil Rights       & 9  & Judicial Power       \\
		3 & First Amendment    & 10 & Federalism           \\
		4 & Due Process        & 11 & Interstate Relations \\
		5 & Privacy            & 12 & Federal Taxation     \\
		6 & Attorneys          & 13 & Miscellaneous        \\
		7 & Unions             & 14 & Private Action      
	\end{tabular}
	\caption{Assigned numbers for the variable \textit{Issue Area}. This information originates from the Supreme Court Database code book.}
\end{table}

@@

\singlespacing
\bibliography{bib} 
\bibliographystyle{apsr}

\onehalfspacing
\appendix
\section{cERGM Estimation}

The normalizing constant in Equation 1 of the main text is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure 4 of the main text---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the cERGM.

The common alternative relies on Monte Carlo methods to approximate the normalizing constant by simulating a large set of networks \citep{hunter2006inference,hummel2012improving}. The resulting estimator, the Monte Carlo MLE (MCMLE), is approximately consistent, meaning that it converges to the MLE as the sample size, i.e., the number of simulated networks, increases. However, one drawback is that with the number of nodes in the Supreme Court citation network being in the order of $10,000$, obtaining the MCMLE is computationally expensive \citep{schmid2017exponential} and the success of the algorithm heavily relies on the starting parameter vector $\theta_0$, which is ideally chosen in the proximity of the unknown MLE \citep{hummel2012improving}. The prevailing choice for $\theta_0$ is the maximum pseudo-likelihood estimation (MPLE) \citep{strauss1990pseudolikelihood}, a fast estimation method that is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated in the main text, assume a logit form. The MPLE is simple to obtain, but does not guarantee a starting value close to the MLE \citep{SchmidHunter2020}.

%We estimate the MCMLE for every term of the cERGM between 1950 and 2015. While the network for the 1950 term consists of $1962$ nodes, the size of the network monotonically increases to $10,020$ nodes for the 2015 term, making estimation more challenging as the network grows. The growth in the network also serves as our reason for estimating a separate model for each term. First, the network is large, with each term introducing the potential for hundreds-of-thousands of new citations. This means we have the statistical power necessary to estimate a separate set of parameter values for each term. Second, based on the findings of \citet{shalizi2013consistency}, we know that the set of parameter values that best fits a network should change as more nodes are added to the network. This means it would be inappropriate to assume a constant set of parameter values as more cases are added each term.

The MCMLE of networks up until the 90s was obtainable in a reasonable time frame starting at the MPLE and sampling $10,000$ networks to approximate the normalizing constant. However, the estimation of most networks in the 90s with the MPLE as starting values was not feasible in an reasonable time frame anymore. Instead, we improved the choice of starting value $\theta_0$ by fixing it at the MCMLE of the previous term $t-1$ and successfully obtained the MCMLE of the network at term $t$. But even this approach started to fail for networks around the turn of the millennium. Neither the MPLE nor the MCMLE of previous terms as starting values led to successful estimation, and neither did the Stepping algorithm \citep{hummel2012improving}. 
The MCMLE for these large citation networks was obtained by setting the starting value according to an novel approach introduced by \citet{SchmidHunter2020}. This method is based on the fact that the MLE of exponential family distributions is solely a function of the vector of sufficient statistics $h(C_t, C_{<t})$, meaning that the MLE of two networks $A$ and $B$ is equal if $h(A)=h(B)$. However, the MPLE of networks with the same sufficient statistics is not necessarily the same. Instead of starting the MCMLE algorithm at the network's MPLE, \citet{SchmidHunter2020} propose searching for a new network $C_t^*$ on the same nodes as the observed network that satisfies $h(C_t, C_{<t}) = h(C_t^*, C_{<t})$, and has a weak dependence structure among unfixed ties. Such a network can be found using simulated annealing algorithms \citep{Kirkpatrick83}. For networks with a weak dependence structure among unfixed ties, the MPLE is similar to the MLE, in addition, the same sufficient statistic between $C_t^*$ and the observed network $C_t$ guarantees the same MLE between these two networks. This makes the MPLE of $C_t^*$ an effective starting value for the MCMLE algorithm. Since for some networks the MCMLE was only obtainable using simulated annealing method to find a starting value, the final results in the paper have all been estimated using simulated annealing.
The simulated annealing algorithm for finding an improved starting value for cERGMs was implemented in the \textbf{cERGM}-package for \textbf{R} \citep{RCore} and can be found at \url{http://github.com/schmid86/cERGM}.

\section{Goodness-of-Fit}
We evaluate the goodness-of-fit of the model following \citet{hunter2008goodness} by examining the distribution of four hyper statistics, e.g., the out- and indegree distribution and the distribution for two different edgewise shared partners statistics. OTP stands for \textit{outgoing two-paths} and refers to the number of cases $r$ that are cited by case $i$ and that cite case $j$, while $j$ is also directly cited by $i$. The second ESP statistic is the OSP specification that has been introduced in section 4.1.2 in the paper. Figure \ref{GOF} visualizes the goodness-of-fit results for the citation network for the 1950 (top) and 2015 (bottom) term. The solid black line indicates the statistic's distribution in the Supreme Court citation network of that given term and the boxplots depict the statistic's distribution of $1000$ networks that have been simulated from the ERGM defined by the MCMLE. This means that in the ideal case the solid black line passes through ever single boxplot.

\begin{figure}[bt]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{GOF_1950}\\
		%\caption{Goodness-of-fit diagnostic for the 1950 network.}
		% \label{GOF_1950}
		\vspace{1cm}
		\includegraphics[width=0.9\textwidth]{GOF_2015}
		\vspace{0.1cm}
		\caption{Goodness-of-fit diagnostic for the 1950 network (top) and the 2015 network (bottom).}
		\label{GOF}
		\vspace{-.25cm}
	\end{center}
\end{figure}

We see that our models do a good job capturing the out and indegree distribution of the citation network, since the black line falls almost exclusively within the ranges spanned by the boxplots. For the ESP distributions we can observe that the number of ties with $r=0$ shared partners is captured well for both the OTP as well as for the OSP statistic. However, the model overestimates the number of $r=1$ shared partners and then, especially in the 2015 term network, underestimates the number of ties with more than $r=1$ shared partners.

\section{Checking for Model Degeneracy}

A common challenge when fitting ERGMs is model degeneracy. Model degeneracy occurs when the probability distribution defined by the parameter vector does not predominantly yield networks with similar statistics as the observed network. Generally, model degeneracy results in simulated networks with no ties or all possible ties. In a non-degenerate model the statistics of the networks that were simulated from the probability distribution defined by the MCMLE fall in the proximity of the observed network's statistics. Figures \ref{mcmcdiagnostics_1950} and \ref{mcmcdiagnostics_2015} depict trace and density plots for the dependence terms in the 1950 and 2015 term citation network. The histograms on the left visualize a statistic's density from $1000$ simulated networks, while the right side shows the statistic's trace plot of the same $1000$ networks. The solid black line indicates the statistic's value in the actual citation network. Both figures indicate that this model is non-degenerate and that the simulated network's statistics fall almost evenly around the observed statistic. The density and trace plots for the ERGM of the terms not depicted provide similar results.

\begin{figure}[bt]
	\begin{center}
		\includegraphics[width=1\textwidth]{Deg_1950}
		\caption{Density and trace plots for the dependency terms of the 1950 term citation network.}
		\label{mcmcdiagnostics_1950}
		\vspace{-.25cm}
	\end{center}
\end{figure}

\begin{figure}[bt]
	\begin{center}
		\includegraphics[width=1\textwidth]{Deg_2015}
		\caption{Density and trace plots for the dependency terms of the 2015 term citation network.}
		\label{mcmcdiagnostics_2015}
		\vspace{-.25cm}
	\end{center}
\end{figure}

\end{document}
